An introduction to Docker for reproducible research

### 2014-11-04 docker-paper/manuscript.Rmd
addressed some of the minor edits from reviewers, see issue #1

---
title: "An introduction to Docker for reproducible research, with examples from the R environment"
author: 
  - name: Carl Boettiger
    affiliation:
      l1: Center for Stock Assessment Research, 
      l2: 110 Shaffer Rd, Santa Cruz, CA 95050, USA 
    email: cboettig(at)gmail.com
    footnote: Corresponding author
abstract: |
  As computational work becomes more and more integral to many aspects of
  scientific research, computational reproducibility has become an issue
  of increasing importance to computer systems researchers and domain
  scientists alike. Though computational reproducibility seems more
  straight forward than replicating physical experiments, the complex
  and rapidly changing nature of computer environments makes being able
  to reproduce and extend such work a serious challenge.  In this paper,
  I explore common reasons that code developed for one research project
  cannot be successfully executed or extended by subsequent researchers.
  I review current approaches to these issues, including virtual machines
  and workflow systems, and their limitations. I then examine how the
  popular emerging technology Docker combines several areas from systems
  research - such as operating system virtualization, cross-platform
  portability, modular re-usable elements, versioning, and a 'DevOps'
  philosophy, to address these challenges.  I illustrate this with several
  examples of Docker use with a focus on the R statistical environment.

@@

<!--
There is increasing interest in promoting repeatable and reproducible research
in computer science.  A community gains confidence in scientific claims when
the experiments supporting those claims are published, examined, and repeated.
Repeating research also helps a community expand on previous results and
discover new directions for investigation.  Repetition requires that the
artifacts used in an experiment be made available to, and made usable by,
investigators.  In other words, artifacts must be shared.  Possible artifacts
include not only binaries, source code, and datasets but also computing and
networking infrastructure such as testbeds and clouds.

@@

Systems research has long concerned itself with the issues of
computational reproducibility and the technologies that can facilitate
those objectives [_e.g._ @Eide_2010, @Harji_2013]. Docker is a new but already
very popular open source tool that combines many of these approaches
in a user friendly implementation, including: (1) performing Linux
container (LXC) based operating system (OS) level virtualization, (2)
portable deployment of containers across platforms, (3) component reuse,
(4) sharing, (5) archiving, and (6) versioning of container images. While
Docker's market success has largely focused on the needs of businesses
in deploying web applications and the potential for a lightweight
alternative to full virtualization, these features have potentially
important implications for systems research in the area of scientific
reproducibility. In this paper, I seek to set these issues in the context
of reproducibility throughout the various domain sciences where 
computation plays an ever-increasing role, but where researchers are
largely unaware of the concerns or technologies involved in making these
computations more reproducible, extensible, and portable to other researchers. 
In so doing, I highlight elements of the Docker platform that should be
of interest to both computer systems research and domain scientists.

<!-- Map here! -->

@@

Critics of the use of VMs to support reproducibility highlight that the 
approach is too much of a black box and thus ill suited for reproducibility [@Watson_2012].
While the approach sidesteps the need to either install or even document the
dependencies, this also makes it more difficult for other researchers to
understand, evaluate, or alter those dependencies.  Moreover, other research
cannot easily build on the virtual machine in a consistent and scalable way. 
If each study provided it's own virtual machine, any pipeline combining the 
tools of multiple studies would quickly become impractical or impossible 
to implement.

@@

Though Docker is not immune to these challenges, it offers a interesting 
example of a way forward in addressing these fundamental concerns. Here 
I highlight five of these features in turn:

@@

First, Docker facilitates modular reuse by build one container on top of
another.  Rather than copy a Dockerfile and then start adding more lines
to the bottom, one can declare a new Dockerfile is built on an old one
using the `FROM` directive. Here is an example Dockerfile that adds the
R statistical computing environment to a basic Ubuntu Linux distribution.

@@

Using Docker containers to distribute reproducible research should be
seen as an approach that is synergistic with, rather than an alternative
to, other technical tools for ensuring computational reproducibility.
Existing tools for managing dependencies for a particular language [_e.g._ R
packages, Ruby gems, etc, see @Ooms_2014] can easily be employed within a
Docker-based approach, allowing the operating-systems level virtualization
to sidestep potential issues such as external library dependencies or
conflicts with existing user libraries.  Other approaches that facilitate
reproducible research also introduce additional software dependencies
and possible points of failure [@FitzJohn_2014]. One example includes
dynamic documents [_e.g._ @Leisch_2002; @Peng_2011; @Xie_2013] which embed
the code required to re-generate the results within the manuscript.  As a
result, it is necessary to package the appropriate typesetting libaries
(_e.g._ \LaTeX) along with the code libaries such that the document executes
successfully for different researchers and platforms.

@@

I noted at the outset that cultural expectations responsible for a
lack of code sharing practices in many fields are a far more extensive
primary barrier to reproducibility than the technical barriers
discussed here. Nevertheless, it may be worth considering how solutions
to these technical barriers can influence the cultural landscape as well.
Many researchers may be reluctant to publish code today because they
fear a it will be primarily a one-way street: more technical savvy
researchers then themselves can benefit from their hard work, while
they may not benefit from the work produced by others.  Lowering the
technical barriers to reuse provides immediate practical benefits
that make this exchange into a more balanced, two-way street. Another
concern is that the difficulty imposed in preparing code to be shared,
such as providing even semi-adequate documentation or support for other
users to be able to install and run it in the first place is too high
[@Stodden_2010]. Thus, lowering these barriers to re-use through
the appropriate infrastructure may also reduce certain cultural
barriers to sharing.

>>>>

---
title: "An introduction to Docker for reproducible research"
author: 
  - name: Carl Boettiger
    affiliation:
      l1: Center for Stock Assessment Research, 
      l2: 110 Shaffer Rd, Santa Cruz, CA 95050, USA 
    email: cboettig(at)gmail.com
    footnote: Corresponding author
abstract: |
  As computational work becomes more and more integral to many aspects of
  scientific research, computational reproducibility has become an issue
  of increasing importance to computer systems researchers and domain
  scientists alike. Though computational reproducibility seems more
  straight forward than replicating physical experiments, the complex
  and rapidly changing nature of computer environments makes being able
  to reproduce and extend such work a serious challenge.  In this paper,
  I explore common reasons that code developed for one research project
  cannot be successfully executed or extended by subsequent researchers.
  I review current approaches to these issues, including virtual machines
  and workflow systems, and their limitations. I then examine how the
  popular emerging technology Docker combines several areas from systems
  research - such as operating system virtualization, cross-platform
  portability, modular re-usable elements, versioning, and a 'DevOps'
  philosophy, to address these challenges.  I illustrate this with several
  examples of Docker use with a focus on the R statistical environment.

@@

<!-- Excerpt from the CALL FOR PAPERS:
There is increasing interest in promoting repeatable and reproducible research
in computer science.  A community gains confidence in scientific claims when
the experiments supporting those claims are published, examined, and repeated.
Repeating research also helps a community expand on previous results and
discover new directions for investigation.  Repetition requires that the
artifacts used in an experiment be made available to, and made usable by,
investigators.  In other words, artifacts must be shared.  Possible artifacts
include not only binaries, source code, and datasets but also computing and
networking infrastructure such as testbeds and clouds.

@@

Systems research has long concerned itself with the issues of
computational reproducibility and the technologies that can facilitate
those objectives [_e.g._ @Eide_2010, @Harji_2013]. Docker is a new but already
very popular open source tool that combines many of these approaches
in a user friendly implementation, including: (1) performing Linux
container (LXC) based operating system (OS) level virtualization, (2)
portable deployment of containers across platforms, (3) component reuse,
(4) sharing, (5) archiving, and (6) versioning of container images. While
Docker's market success has largely focused on the needs of businesses
in deploying web applications and the potential for a lightweight
alternative to full virtualization, these features have potentially
important implications for systems research in the area of scientific
reproducibility.

In this paper, I seek to address two audiences. First, that of the domain
scientist, those conducting research in ecology, bioinformatics, economics,
psychology and so many other disciplines in which computation plays an 
ever-increasing role. I seek to help this audience become more aware of the 
concerns and challenges in making these computations more reproducible, extensible, 
and portable to other researchers, and highlight Docker as platform to address 
these challenges. The second audience is that of the computer systems researchers:
readers familiar with both the challenges to reproducibility (versioning,
software dependencies, etc.) and the technical components underlying Docker (virtualization,
LXC containers, jails, hashes, etc.), but who may be unfamiliar with Docker software. 
I hope the domain scientist will be motivated to try Docker as a tool to address
reproducibility. Meanwhile, I expect the systems researcher to see Docker technology as
an area ripe for further systems research (e.g. to what extent can differences between 
Linux kernels frustrate reproducibility in a Docker image?) rather than as a tool
for their own study (because Docker applies only to software at the level above
the Linux kernel, it will be of little use in making studies of differences in 
kernel behavior, hardware performance, execution times, and so forth reproducible.)

@@

Critics of the use of VMs to support reproducibility highlight that the 
approach is too much of a black box and thus ill suited for reproducibility [@Watson_2012].
While the approach sidesteps the need to either install or even document the
dependencies, this also makes it more difficult for other researchers to
understand, evaluate, or alter those dependencies.  Moreover, other research
cannot easily build on the virtual machine in a consistent and scalable way. 
If each study provided its own virtual machine, any pipeline combining the 
tools of multiple studies would quickly become impractical or impossible 
to implement.

@@

Though Docker is not immune to these challenges, it offers an interesting 
example of a way forward in addressing these fundamental concerns. Here 
I highlight five of these features in turn:

@@

First, Docker facilitates modular reuse by building one container on top of
another.  Rather than copy a Dockerfile and then start adding more lines
to the bottom, one can declare a new Dockerfile is built on an old one
using the `FROM` directive. Here is an example Dockerfile that adds the
R statistical computing environment to a basic Ubuntu Linux distribution.

@@

Using Docker containers to distribute reproducible research should be
seen as an approach that is synergistic with, rather than an alternative
to, other technical tools for ensuring computational reproducibility.
Existing tools for managing dependencies for a particular language [_e.g._ R
packages, Ruby gems, etc, see @Ooms_2014] can easily be employed within a
Docker-based approach, allowing the operating-systems level virtualization
to sidestep potential issues such as external library dependencies or
conflicts with existing user libraries.  Other approaches that facilitate
reproducible research also introduce additional software dependencies
and possible points of failure [@FitzJohn_2014]. One example includes
dynamic documents [_e.g._ @Leisch_2002; @Peng_2011; @Xie_2013] which embed
the code required to re-generate the results within the manuscript.  As a
result, it is necessary to package the appropriate typesetting libraries
(_e.g._ \LaTeX) along with the code libraries such that the document executes
successfully for different researchers and platforms.

@@

I noted at the outset that cultural expectations responsible for a
lack of code sharing practices in many fields are a far more extensive
primary barrier to reproducibility than the technical barriers
discussed here. Nevertheless, it may be worth considering how solutions
to these technical barriers can influence the cultural landscape as well.
Many researchers may be reluctant to publish code today because they
fear a it will be primarily a one-way street: more technical savvy
researchers than themselves can benefit from their hard work, while
they may not benefit from the work produced by others.  Lowering the
technical barriers to reuse provides immediate practical benefits
that make this exchange into a more balanced, two-way street. Another
concern is that the difficulty imposed in preparing code to be shared,
such as providing even semi-adequate documentation or support for other
users to be able to install and run it in the first place is too high
[@Stodden_2010]. Thus, lowering these barriers to re-use through
the appropriate infrastructure may also reduce certain cultural
barriers to sharing.
